Towards Universal Forest Mapping: Foundation Embeddings Outperform Hand-Crafted Harmonics for Deciduous–Evergreen Classification

### 2025-10-31 article/article.tex
safe save

\begin{abstract}
High-resolution, annually updated deciduous–evergreen maps underpin forest carbon accounting, disturbance surveillance, and biodiversity reporting, yet most agencies still lack a practical way to produce them. Building such maps through mathematical simulation over remote sensing time series typically means choosing between hand-crafted harmonic descriptors that demand local phenology expertise and deep models that require GPU fleets and machine-learning specialists—the ``compute-or-craft'' bottleneck. Unfortunately, neither option removes the bottleneck for resource-limited agencies. We test whether pre-computed AlphaEarth embeddings, distributed through Google Earth Engine, can remove that constraint when evaluated fairly against physics-informed harmonics. Using 14.1 million labelled pixels grouped into 639 GRECO-balanced tiles, we train identical eco-region-weighted Random Forests on two 14-dimensional feature sets: a fold-selected subset of AlphaEarth embeddings (EMB-14) and Sentinel-2 harmonic descriptors (HARM-14). EMB-14 increases overall accuracy from $0.904 \pm 0.0059$ to $0.926 \pm 0.0059$, lifts macro-F1 from 0.874 to 0.905, halves the expected calibration error (0.033 vs 0.059), and reduces national edge and patch densities by 59\,\% and 68\,\%, while a frozen 2023 model transfers to 2018–2022 embeddings with macro IoU ranging from 0.786 to 0.801. Because the strongest performance now arrives without bespoke compute or bespoke features, label coverage becomes the primary lever for improving nationwide deciduous–evergreen maps.

\keyFont{\section{Keywords:} deciduous-evergreen, AlphaEarth embeddings, harmonic analysis, Random Forest, phenology}
\end{abstract}

@@

High-resolution, annually refreshed deciduous–evergreen maps are essential for tracking carbon fluxes, prioritising disturbance response, and reporting biodiversity trends \citep{Zhu2014,Zhao2019,Verbesselt2010a,Verbesselt2010b,Kennedy2010,Kennedy2018}. Yet most agencies still rely on legacy inventories or continental products that update slowly and blur key forest types. The Copernicus Dominant Leaf Type (DLT) map, for instance, separates broadleaf from conifer but merges evergreen oaks with pines \citep{EU2024a}, while BD Forêt V2 offers detailed stand labels frozen to field campaigns from 2007–2018 \citep{IGN2024}. Delivering a current nationwide map therefore remains a bespoke effort rather than an operational service.

Those bespoke efforts confront the same dilemma: build hand-crafted features (labour-intensive) or train deep networks (compute-intensive). Physics-informed workflows fit harmonic curves to Sentinel-2 indices, extracting amplitudes, phases, and residual variances that capture seasonal behaviour \citep{Inglada2017,Li2023,Bolton2020}. The descriptors are interpretable and appealing to domain experts, but the tuning effort scales with every eco-region, sensor update, and disturbance regime. In parallel, end-to-end deep networks can learn spatial context automatically \citep{Low2020,Xie2024FoundationEffective}, yet they demand GPU clusters, large labelled datasets, and dedicated ML operations. This compute-or-craft bottleneck still limits operational deployment of high-resolution forest mapping.

Foundation embeddings offer a potential way out by shifting the constraint: pre-trained models published through Earth Engine eliminate both GPU clusters and fragile harmonic tuning, concentrating effort on the one input that cannot be outsourced—ground-truth labels. Satellite pretraining initiatives such as SatMAE, Prithvi-EO, and AlphaEarth learn multi-sensor, multi-temporal representations that transfer across tasks \citep{Cong2022,Szwarcman2024PrithviEO2,AlphaEarth2025}. AlphaEarth is particularly compelling for practitioners because it publishes annual 64-dimensional embeddings at 10\,m directly through Google Earth Engine—no GPU, no feature tuning required. These vectors have already supported cross-border vegetation typing and disturbance analyses \citep{Houriez2025AEFDataGen,Seydi2025AlphaEarthBurnedArea}, but their value relative to carefully engineered harmonics has not been benchmarked under matched conditions. Figure~\ref{fig:teaser_sat_emb_class} previews the contrast: embeddings sharpen phenological structure without bespoke tuning.

We address that gap with a France-wide benchmark that evaluates harmonics and embeddings on equal footing. Our dataset aggregates 14.1 million labelled forest pixels from four sources, groups them into 639 non-overlapping 2.5\,km tiles, and balances sampling across the eleven GRECO eco-regions. Two compact 14-dimensional feature sets feed identical eco-region-weighted Random Forests: the fold-selected AlphaEarth subset (\textbf{EMB-14}) and physics-informed Sentinel-2 harmonic descriptors (\textbf{HARM-14}). Tile-grouped cross-validation measures accuracy and calibration, ancillary drivers explain spatial heterogeneity, temporal experiments test whether a frozen 2023 embedding model transfers to 2018–2022 vintages, and national comparisons situate the outputs against Copernicus DLT and BD Forêt.

This design lets us interrogate three hypotheses. \textbf{H1} asks whether embeddings match or exceed harmonic accuracy and calibration when data, folds, and classifiers are held constant. \textbf{H2} examines whether embeddings deliver smoother maps whose gains co-vary with soils, climate, and forest composition. \textbf{H3} tests temporal robustness and compatibility with legacy national products. The remainder of the paper details the data and processing pipeline (Section~2), presents the accuracy, coherence, ancillary, and temporal analyses (Section~3), and concludes with operational guidance and remaining limitations (Section~4).

@@

\subsection{Study area and eco-regional stratification}
\label{sec:greco}
We focus on mainland France and Corsica, where temperate, montane, and Mediterranean forest types co-exist within short distances. To preserve ecological diversity in both training and evaluation we rely on the eleven \emph{Grandes Régions Écologiques} (GRECO) defined by the French National Forest Inventory \citep{IGN2013GRECO}. These eco-regions capture the dominant gradients from Atlantic mixed forests through semi-continental plateaus to Mediterranean evergreen stands and Alpine massifs. We partitioned the national forest mask into 639 non-overlapping 2.5\,km\,$\times$\,2.5\,km tiles (Figure~\ref{fig:training_tiles}), retaining tiles that contain at least ten inventory or plot-labelled pixels and are separated by 5\,km to limit spatial autocorrelation. After the reference quota is met, remaining forest pixels in each tile are populated with buffered BD Forêt V2 polygons so that both feature families share an identical footprint. Sample weights then align the dataset with the actual forest extent of each eco-region. We estimate that extent through an effective forest area \(A_r\), obtained by multiplying each GRECO polygon by the region's mean forest cover ratio provided by the National Forest Inventory, and we sum these areas across regions to get \(\sum_{j} A_j\). The training dataset contains \(N\) labelled forest pixels, with \(n_r\) of them belonging to region \(r\); the per-pixel weight is therefore defined as \(w_r = (A_r / \sum_{j} A_j) / (n_r / N)\). Regions that occupy a large share of forest but are sparse in the sample receive \(w_r > 1\), whereas regions that are already over-represented are down-weighted. For example, the Oceanic Southwest accounts for roughly 22\% of the effective forest area yet only about 21\% of the labelled pixels, so its observations are scaled by \(w_r \approx 1.06\). We normalise the weights so that \(\sum_i w_i = N\); the resulting weights range from 0.89 in densely sampled regions to 1.48 in rare mountainous tiles. A small fraction of coastal pixels remains outside the GRECO mask and is excluded from stratified summaries.

The GRECO framework offers concise descriptors that explain the spatial behaviour of our models. Region A (Greater Crystalline and Oceanic West) is a humid Atlantic bocage of oak–chestnut forests interspersed with post-war conifer plantations and dense hedgerows. Region B (Semi-Oceanic North Centre) covers the Paris Basin cuestas with extensive pedunculate/sessile oak and hornbeam on loess plateaus and Scots pine on sandy soils. Region C (Greater Semi-Continental East) spans the Ardennes and Lorraine, with beech–fir forests on cool uplands. Region D (Vosges) combines silver fir–beech on steep crystalline slopes, while Region E (Jura) hosts calcareous beech–fir mosaics above karst plateaus. Region F (Oceanic Southwest) includes the Landes maritime pine massif and humid mixed stands. Region G (Central Massif) contains montane beech–fir and chestnut belts, Region H (Alps) sharply transitions from deciduous foothills to coniferous subalpine belts, and Region I (Pyrenees) alternates Atlantic and Mediterranean influences over steep gradients. Region J (Mediterranean) is dominated by evergreen holm and cork oak maquis with summer drought stress, and Region K (Corsica) blends lowland maquis with high-elevation Laricio pine forests. These descriptors provide the ecological backdrop for the regional analyses reported in Section~\ref{sec:results} and are summarised in Table~\ref{tab:greco_summary}.

@@

\begin{figure}[H]
    \centering
    \includegraphics[width=0.8\textwidth]{images/tiles_2_5_km_final_visualization.png}
    \caption{Distribution of the 639 eco-region balanced training tiles. Warm colours indicate Mediterranean and montane eco-regions, while cool colours denote Atlantic and semi-continental domains. Each tile contains 10\,m pixels labelled as deciduous or evergreen and is used as an indivisible unit during cross-validation.}
    \label{fig:training_tiles}
\end{figure}

@@

\subsection{Feature extraction}
\subsubsection{Sentinel-2 harmonic descriptors}
We reconstruct Sentinel-2 Level-2A surface reflectances from Google Earth Engine monthly MEDIAN composites, applying the s2cloudless cloud-probability mask with a 75\,\% threshold, removing QA60 bits 10 and 11 (opaque and cirrus clouds), and discarding scenes with more than 95\,\% cloudy pixels. No gap filling or scene classification layer is applied. For each tile we derive the vegetation indices NDVI, EVI, NBR, and the SWIR ratio (CRSWIR). The annual signal of each index is fitted with two sinusoidal harmonics through ordinary least squares,
\[
  x(t) = C + \sum_{k=1}^{2} \big[a_k \cos\!\big(2\pi k t/T\big) + b_k \sin\!\big(2\pi k t/T\big)\big], \quad T = 1\,\text{year}.
\]
We recover offsets, amplitudes, phases, and residual variance; phases are transformed into their sine and cosine components to avoid wrap-around discontinuities. Figure~\ref{fig:harmonics_decomposition} illustrates how the two-harmonic model decomposes an NDVI trajectory into interpretable amplitude and phase components while keeping residuals explicit. Recursive feature elimination (Section~\ref{subsubsec:rfe}) retained 14 descriptors that balance interpretability and cross-validated accuracy (Supplementary Section~S1). These descriptors (Table~\ref{tab:harmonic14}) span amplitude, phase, offset, and residual components across four indices: NDVI first-harmonic amplitude captures seasonal greenness pulses; NBR amplitude reflects moisture/structural contrast; CRSWIR offset encodes baseline water content. We denote this subset as \textbf{HARM-14}.

\begin{table}[H]
    \centering
    \small
    \begin{tabular}{lp{0.62\textwidth}}
        \hline
        \textbf{HARM-14 descriptor} & \textbf{Ecological signal} \\
        \hline
        NDVI first-harmonic amplitude & Seasonal strength of broadleaf greenness pulses \\
        NDVI first-harmonic phase (cosine/sine) & Timing of green-up and senescence transitions \\
        NDVI second-harmonic phase (sine) & Asymmetry between rapid spring and gradual autumn trajectories \\
        NDVI offset & Mean canopy greenness throughout the year \\
        NBR first-harmonic amplitude & Annual moisture and structural contrast between canopies and soil \\
        NBR first-harmonic phase (cosine) & Calendar timing of minimum fuel moisture \\
        NBR second-harmonic phase (cosine) & Secondary moisture cycle in bimodal climates \\
        NBR offset & Average woody biomass signal \\
        NBR residual variance & Short-term disturbance departures from harmonic behaviour \\
        CRSWIR first-harmonic phase (cosine) & Timing of peak shortwave-infrared water stress \\
        CRSWIR second-harmonic phase (cosine) & Recovery pattern of evergreen water content \\
        CRSWIR offset & Mean canopy water and lignin content \\
        CRSWIR residual variance & Fine-scale heterogeneity in SWIR response \\
        \hline
    \end{tabular}
    \caption{Fourteen harmonic descriptors retained after eco-region-balanced recursive feature elimination. Together they define the \textbf{HARM-14} feature set used in the Random Forest baseline. Phases are represented via sine and cosine components to avoid angular discontinuities.}
    \label{tab:harmonic14}
\end{table}

\begin{figure}[H]
    \centering
    \includegraphics[width=\textwidth]{images/harmonics_decomposition.png}
    \caption{Harmonic decomposition of an NDVI annual curve. The first and second harmonics capture the asymmetric spring green-up and autumn senescence, while the residual component highlights departures linked to disturbance or mixed pixels. These interpretable pieces constitute the descriptors retained in \textbf{HARM-14}.}
    \label{fig:harmonics_decomposition}
\end{figure}

\subsubsection{AlphaEarth embeddings}
AlphaEarth provides annual 64-dimensional embeddings at 10\,m resolution, learned from multi-sensor stacks (Sentinel-1/2 optical reflectances, Sentinel-1 backscatter, Landsat seasonal composites, GEDI structure, ERA5-Land climate predictors) by masking and reconstructing spatio-temporal tokens \citep{AlphaEarth2025}. The resulting vector can be viewed as a "spectro-temporal fingerprint" of each pixel: it summarises how the site looks across the year, how it relates to its neighbourhood, and how ancillary drivers (elevation, climate) modulate that pattern. Because the embeddings are already computed and distributed through Earth Engine, practitioners can treat them as off-the-shelf features without training deep networks. We retrieved the 2018, 2020, 2022, and 2023 embedding rasters, converted the tiles to Parquet, and removed pixels with non-finite components (mostly coastline artifacts). To align with the harmonic dataset we perform an inner join on tile, row, and column indices so that both feature families use the exact same labelled pixels and eco-region weights. The 14 embedding dimensions retained after feature elimination (Section~\ref{subsubsec:rfe}) define the \textbf{EMB-14} feature set used in training and evaluation (Table~\ref{tab:emb14}).

\begin{table}[H]
    \centering
    \small
    \begin{tabular}{ll}
        \hline
        \textbf{Rank} & \textbf{EMB-14 dimension} \\
        \hline
        1 & embedding\_46 \\
        2 & embedding\_18 \\
        3 & embedding\_5 \\
        4 & embedding\_30 \\
        5 & embedding\_39 \\
        6 & embedding\_0 \\
        7 & embedding\_57 \\
        8 & embedding\_23 \\
        9 & embedding\_6 \\
        10 & embedding\_15 \\
        11 & embedding\_13 \\
        12 & embedding\_22 \\
        13 & embedding\_11 \\
        14 & embedding\_24 \\
        \hline
    \end{tabular}
    \caption{Fourteen AlphaEarth embedding dimensions retained after eco-region-balanced recursive feature elimination, ranked by mean importance across folds. Together they define the \textbf{EMB-14} feature set.}
    \label{tab:emb14}
\end{table}

\subsubsection{Recursive feature elimination}
\label{subsubsec:rfe}
Both feature families are pruned with the same eco-region-balanced recursive elimination. Starting from the full candidate list (all harmonic descriptors after circular transforms or the 64 embedding channels), we train weighted random forests on each of the five spatial folds, rank features by their mean importance, and remove the weakest features at every stage. The elimination schedule removes features successively, ensuring that at least 10 predictors remain. After every pruning step we record macro-F1 and accuracy along with eco-region metrics; the curve plateaus at 14 features for the harmonics. We therefore retain the best-performing 14-feature configuration and refer to it as HARM-14 for harmonics and EMB-14 for embeddings. Alternative top-$K$ lists extracted from the same procedure support the ablation studies summarised in Supplementary Section~S2.

\subsection{Random Forest training and cross-validation}
We train separate random forest classifiers for HARM-14 and EMB-14 using 50 trees, maximum depth 30, minimum samples split 30, and minimum leaf size 15. These hyperparameters come from scikit-learn’s successive halving grid search: candidate configurations are evaluated on progressively larger sample fractions while eliminating the weakest options at each stage, yielding a coarse-to-fine search that is more sample-efficient than an exhaustive grid. We tune the search on the harmonic features and reuse the winning configuration for the embedding model so that any differences stem from the representation rather than model capacity—a choice that slightly favours the harmonic baseline and therefore provides a conservative comparison for EMB-14. Class imbalance is mitigated through scikit-learn’s balanced class weights combined with the eco-region weights described above. Five-fold cross-validation preserves tile integrity: tiles within each eco-region are grouped and assigned wholesale to folds, ensuring that no validation tile shares pixels with the training subset. The folds are identical for both feature sets; we record out-of-fold predictions, per-fold metrics, and the aggregated confusion matrix. After cross-validation we retrain each model on the full weighted dataset for downstream inference.

>>>>

\begin{abstract}
Forest agencies need annual deciduous–evergreen maps to monitor carbon budgets, disturbance recovery, and biodiversity compliance, yet most lack a workflow that scales beyond bespoke campaigns. Existing options demand a compute-or-craft trade-off: either engineer harmonic descriptors with local phenology expertise or train deep networks on GPU fleets.
We evaluate pre-computed AlphaEarth embeddings—the state-of-the-art foundation model now hosted on Google Earth Engine—as a third path that keeps operations lightweight while encoding neighbourhood context that temporal harmonics miss. Using 14.1 million labelled pixels grouped into 639 eco-region-balanced tiles, we train logistic regression, linear support-vector machines, and eco-region-weighted Random Forests on matched 14-dimensional feature sets: physics-informed harmonics (HARM) and AlphaEarth embeddings (EMB).
Across the three learners, EMB lifts accuracy by $2.8 \pm 0.4$ percentage points, macro-F1 by $3.7 \pm 0.5$, halves expected calibration error (0.033 vs 0.059), and reduces national map fragmentation by 68\,\% relative to HARM. A 2023 model transfers to 2018–2022 embeddings with macro IoU between 0.786 and 0.801, showing that the embeddings track inter-annual drift without re-fitting.
By removing bespoke compute and feature engineering, AlphaEarth shifts the bottleneck to label coverage, letting agencies focus on ground truth collection rather than infrastructure.

\keyFont{\section{Keywords:} deciduous-evergreen, AlphaEarth embeddings, harmonic analysis, Random Forest, logistic regression, support vector machine, phenology}
\end{abstract}

@@

High-resolution, annually refreshed deciduous–evergreen maps underpin carbon accounting, disturbance response, and biodiversity reporting \citep{Zhu2014,Zhao2019,Verbesselt2010a,Verbesselt2010b,Kennedy2010,Kennedy2018}. Yet most agencies still rely on legacy inventories or continental products that update slowly and blur key forest types. The Copernicus Dominant Leaf Type map separates broadleaf from conifer but merges evergreen oaks with pines \citep{EU2024a}. BD Forêt V2 provides detailed stand labels yet reflects field campaigns from 2007–2018 \citep{IGN2024}. Delivering a current nationwide map therefore remains a bespoke effort rather than an operational service.

Bespoke projects confront a compute-or-craft dilemma. Physics-informed workflows fit harmonic curves to Sentinel-2 indices, extracting amplitudes, phases, and residual variances that summarise seasonal behaviour \citep{Inglada2017,Li2023,Bolton2020}. The descriptors are interpretable and appealing to domain experts, but the tuning effort scales with every eco-region, sensor update, and disturbance regime.

Deep networks promise automatic feature learning \citep{Low2020,Xie2024FoundationEffective}, yet they demand GPU clusters, large labelled datasets, and machine-learning operations that many agencies cannot maintain. The result is a compute-or-craft bottleneck that still blocks operational deployment.

Foundation embeddings offer a third path by shifting the constraint. Pre-trained models distributed through Google Earth Engine remove both the GPU requirement and the fragility of hand-crafted harmonics, concentrating effort on the one ingredient that cannot be outsourced: ground-truth labels. These embeddings encode spatial neighbourhood context alongside temporal phenology, a capability that traditional harmonic descriptors miss. Figure~\ref{fig:teaser_sat_emb_class} previews how embeddings sharpen structure without manual feature design.

Satellite pretraining initiatives such as SatMAE, Prithvi-EO, and AlphaEarth learn multi-sensor, multi-temporal representations that transfer across tasks \citep{Cong2022,Szwarcman2024PrithviEO2,AlphaEarth2025}. AlphaEarth, a state-of-the-art foundation model, is particularly compelling because it publishes annual 64-dimensional embeddings at 10\,m resolution via Google Earth Engine. The vectors have already supported cross-border vegetation typing and disturbance analyses \citep{Houriez2025AEFDataGen,Seydi2025AlphaEarthBurnedArea}, yet no study has benchmarked them against carefully engineered harmonics under matched conditions.

We address that gap with a France-wide benchmark that evaluates harmonics and embeddings on equal footing. Our dataset aggregates 14.1 million labelled forest pixels from four sources, groups them into 639 non-overlapping 2.5\,km tiles, and balances sampling across the eleven GRECO eco-regions. Two compact 14-dimensional feature sets feed three lightweight classifiers—logistic regression, linear support-vector machines, and eco-region-weighted Random Forests. HARM denotes the physics-informed Sentinel-2 harmonic descriptors, and EMB denotes the fold-selected AlphaEarth subset. Tile-grouped cross-validation measures accuracy and calibration, ancillary drivers explain spatial heterogeneity, temporal experiments test whether a frozen 2023 embedding model transfers to 2018–2022 vintages, and national comparisons situate the outputs against Copernicus DLT and BD Forêt.

We hypothesise that pre-computed embeddings outperform engineered harmonics for deciduous–evergreen mapping while remaining operationally lightweight. We test this claim through accuracy and calibration analyses, spatial coherence and environmental driver assessments, and temporal robustness experiments that include comparisons with national inventories. The remainder of the paper details the data and processing pipeline, reports the results, and concludes with operational guidance and remaining limitations.

@@

\subsection{Study area and eco-regional stratification}
\label{sec:greco}
We require spatially representative training data that span the main forest gradients of mainland France and Corsica. Temperate lowlands, montane belts, and Mediterranean shrublands sit within short distances, so we rely on the eleven \emph{Grandes Régions Écologiques} (GRECO) defined by the French National Forest Inventory \citep{IGN2013GRECO}. The GRECO zones distil climate and forest structure transitions from Atlantic mixed forests through semi-continental plateaus to Mediterranean evergreen stands and Alpine massifs.

We partition the national forest mask into 639 non-overlapping 2.5\,km\,$\times$\,2.5\,km tiles (Figure~\ref{fig:training_tiles}). Each tile contains at least ten inventory or plot-labelled pixels and is separated from its neighbours by 5\,km to limit spatial autocorrelation. Once the reference quota is met, remaining forest pixels in each tile are populated with buffered BD Forêt V2 polygons so that both feature families share an identical footprint.

To keep training and evaluation aligned with the true forest extent we apply eco-region weights. For each region \(r\) we compute an effective forest area \(A_r\) by multiplying the GRECO polygons by the mean forest cover ratio reported by the inventory and summing them to obtain \(\sum_j A_j\). With \(N\) labelled pixels overall and \(n_r\) drawn from region \(r\), the per-pixel weight becomes \(w_r = (A_r / \sum_j A_j) / (n_r / N)\). Regions covering 22\,\% of forests but only 21\,\% of pixels, such as the Oceanic Southwest, are scaled by \(w_r \approx 1.06\); over-represented plains drop to \(w_r \approx 0.89\). We normalise the weights so that \(\sum_i w_i = N\), which keeps rare mountain forests from being drowned out by the plains while avoiding over-sampling common stands. The resulting weights range from 0.89 in densely sampled regions to 1.48 in rare mountainous tiles. A small fraction of coastal pixels lies outside the GRECO mask and is excluded from stratified summaries.

The GRECO framework also provides ecological descriptors that contextualise the spatial behaviour reported in Section~\ref{sec:results}. Atlantic bocage and plantation mosaics dominate Regions A and F, semi-continental oak–beech plateaus span Regions B and C, montane belts extend across the Central Massif, Alps, and Pyrenees (Regions G–I), while drought-prone evergreen maquis defines the Mediterranean mainland and Corsica (Regions J and K). Table~\ref{tab:greco_summary} summarises the climate, forest structure, and training sample counts for each region.

@@

\begin{figure}[H]
    \centering
    \includegraphics[width=0.8\textwidth]{images/tiles_2_5_km_final_visualization.png}
    \caption{Distribution of the 639 eco-region balanced training tiles. Warm colours indicate Mediterranean and montane eco-regions, while cool colours denote Atlantic and semi-continental domains. Each tile contains at least ten labelled pixels, is separated from its neighbours by 5\,km, and serves as an indivisible unit during cross-validation.}
    \label{fig:training_tiles}
\end{figure}

@@

\subsection{Feature extraction}
\subsubsection{Sentinel-2 harmonic descriptors}
We reconstruct Sentinel-2 Level-2A surface reflectances from Google Earth Engine monthly MEDIAN composites, applying the s2cloudless cloud-probability mask with a 75\,\% threshold, removing QA60 bits 10 and 11 (opaque and cirrus clouds), and discarding scenes with more than 95\,\% cloudy pixels. No gap filling or scene classification layer is applied. For each tile we derive the vegetation indices NDVI, EVI, NBR, and the SWIR ratio (CRSWIR). The annual signal of each index is fitted with two sinusoidal harmonics through ordinary least squares,
\[
  x(t) = C + \sum_{k=1}^{2} \big[a_k \cos\!\big(2\pi k t/T\big) + b_k \sin\!\big(2\pi k t/T\big)\big], \quad T = 1\,\text{year}.
\]
We recover offsets, amplitudes, phases, and residual variance; phases are transformed into their sine and cosine components to avoid wrap-around discontinuities. These features are purely temporal: each pixel's time series is modelled independently and ignores the surrounding context. Figure~\ref{fig:harmonics_decomposition} illustrates how the two-harmonic model decomposes an NDVI trajectory into interpretable amplitude and phase components while keeping residuals explicit. Recursive feature elimination (Section~\ref{subsubsec:rfe}) retained 14 descriptors that balance interpretability and cross-validated accuracy (Supplementary Section~S1). These descriptors (Table~\ref{tab:harmonic14}) span amplitude, phase, offset, and residual components across four indices: NDVI first-harmonic amplitude captures seasonal greenness pulses; NBR amplitude reflects moisture and structural contrast; CRSWIR offset encodes baseline water content. We denote this 14-dimensional subset as \textbf{HARM}.

\begin{table}[H]
    \centering
    \small
    \begin{tabular}{lp{0.62\textwidth}}
        \hline
        \textbf{HARM descriptor} & \textbf{Ecological signal} \\
        \hline
        NDVI first-harmonic amplitude & Seasonal strength of broadleaf greenness pulses \\
        NDVI first-harmonic phase (cosine/sine) & Timing of green-up and senescence transitions \\
        NDVI second-harmonic phase (sine) & Asymmetry between rapid spring and gradual autumn trajectories \\
        NDVI offset & Mean canopy greenness throughout the year \\
        NBR first-harmonic amplitude & Annual moisture and structural contrast between canopies and soil \\
        NBR first-harmonic phase (cosine) & Calendar timing of minimum fuel moisture \\
        NBR second-harmonic phase (cosine) & Secondary moisture cycle in bimodal climates \\
        NBR offset & Average woody biomass signal \\
        NBR residual variance & Short-term disturbance departures from harmonic behaviour \\
        CRSWIR first-harmonic phase (cosine) & Timing of peak shortwave-infrared water stress \\
        CRSWIR second-harmonic phase (cosine) & Recovery pattern of evergreen water content \\
        CRSWIR offset & Mean canopy water and lignin content \\
        CRSWIR residual variance & Fine-scale heterogeneity in SWIR response \\
        \hline
    \end{tabular}
    \caption{Fourteen harmonic descriptors retained after eco-region-balanced recursive feature elimination. Together they define the \textbf{HARM} feature set used in the Random Forest baseline. Higher values increase seasonal amplitude or evergreen likelihood depending on the index. Phases are represented via sine and cosine components to avoid angular discontinuities.}
    \label{tab:harmonic14}
\end{table}

\begin{figure}[H]
    \centering
    \includegraphics[width=\textwidth]{images/harmonics_decomposition.png}
    \caption{Harmonic decomposition of an NDVI annual curve. The first and second harmonics capture the asymmetric spring green-up and autumn senescence, while the residual component highlights departures linked to disturbance or mixed pixels. These interpretable pieces constitute the descriptors retained in \textbf{HARM}.}
    \label{fig:harmonics_decomposition}
\end{figure}

\subsubsection{AlphaEarth embeddings}
AlphaEarth provides annual 64-dimensional embeddings at 10\,m resolution, learned from multi-sensor stacks (Sentinel-1/2 optical reflectances, Sentinel-1 backscatter, Landsat seasonal composites, GEDI structure, ERA5-Land climate predictors) by masking and reconstructing spatio-temporal tokens \citep{AlphaEarth2025}. Each embedding pixel therefore encodes how a site looks across the seasons and how it relates to its neighbouring pixels; the pre-training injects spatial context that purely temporal harmonics cannot capture. This neighbourhood awareness helps separate mixed forest patches that follow similar temporal curves but diverge in structure. Because the embeddings are already computed and distributed through Earth Engine, practitioners can treat them as off-the-shelf features without training deep networks. We retrieved the 2018, 2020, 2022, and 2023 embedding rasters, converted the tiles to Parquet, and removed pixels with non-finite components (mostly coastline artifacts). To align with the harmonic dataset we perform an inner join on tile, row, and column indices so that both feature families use the exact same labelled pixels and eco-region weights. The 14 embedding dimensions retained after feature elimination (Section~\ref{subsubsec:rfe}) define the \textbf{EMB} feature set used in training and evaluation (Table~\ref{tab:emb14}); they summarise the dominant spatial–seasonal fingerprints encoded by AlphaEarth.

\begin{table}[H]
    \centering
    \small
    \begin{tabular}{ll}
        \hline
        \textbf{Rank} & \textbf{EMB dimension} \\
        \hline
        1 & embedding\_46 \\
        2 & embedding\_18 \\
        3 & embedding\_5 \\
        4 & embedding\_30 \\
        5 & embedding\_39 \\
        6 & embedding\_0 \\
        7 & embedding\_57 \\
        8 & embedding\_23 \\
        9 & embedding\_6 \\
        10 & embedding\_15 \\
        11 & embedding\_13 \\
        12 & embedding\_22 \\
        13 & embedding\_11 \\
        14 & embedding\_24 \\
        \hline
    \end{tabular}
    \caption{Fourteen AlphaEarth embedding dimensions retained after eco-region-balanced recursive feature elimination, ranked by mean importance across folds. Together they define the \textbf{EMB} feature set. Higher values increase evergreen probability in the fitted Random Forest.}
    \label{tab:emb14}
\end{table}

\subsubsection{Recursive feature elimination}
\label{subsubsec:rfe}
Both feature families are pruned with the same eco-region-balanced recursive elimination. Starting from the full harmonic descriptor list (after circular transforms) or the 64 embedding channels, we train weighted Random Forests on each spatial fold, rank predictors by mean importance, and drop the weakest until performance plateaus. The 14-feature configuration maximises macro-F1 for both representations and defines the HARM and EMB subsets used throughout. Supplementary Section~S1 reports the full elimination schedule and ablations for alternative top-$K$ lists.

\subsection{Classifier training, tuning, and cross-validation}
We evaluate three lightweight classifiers on each feature family: logistic regression, linear support-vector machines (SVM), and Random Forests. This choice keeps modelling simple so that any performance gap reflects feature quality rather than model complexity; if embeddings prevail under these conditions, agencies can deploy them without deep learning infrastructure. Every model is trained with class-balanced sample weights on the same five eco-region-balanced folds described above, ensuring that differences in performance arise from the feature representation rather than from split variability.

Hyperparameters are selected with scikit-learn’s successive halving grid search. For each estimator we define a compact grid, evaluate candidates on progressively larger fractions of the harmonic dataset, and reuse the best configuration when fitting on the embeddings. This harmonics-first tuning strategy slightly favours the handcrafted baseline and therefore yields conservative estimates of the embedding gains. Logistic regression combines \texttt{StandardScaler} with an $\ell_2$ or elastic-net penalty (grid over $C\in\{0.01,0.1,1,10\}$ and $l_1$ ratios $\{0,0.5,0.9\}$) using the \texttt{saga} solver. The linear SVM applies the same scaling and explores hinge versus squared-hinge losses with $C\in\{0.01,0.1,1,10\}$. The Random Forest grid matches the original pipeline (50--100 trees, maximum depth 15--None, minimum split 15--60, minimum leaf 10--20), and the selected configuration uses 50 trees, maximum depth 30, minimum split 30, and minimum leaf 15. Logistic regression and the Random Forest provide calibrated probabilities used later for reliability analysis; the linear SVM contributes decision scores only. After hyperparameter selection, each model is retrained on the full weighted dataset (harmonics or embeddings) and the cross-validation artefacts—fold metrics, out-of-fold predictions, and fitted models—are archived for subsequent analysis.

\subsection{Evaluation metrics and derived summaries}
\label{sec:metrics}
Fold-level metrics include overall accuracy, class-specific precision and recall, macro- and weighted-F1, and the four entries of the confusion matrix. Macro-F1 averages the class-specific F1-scores with equal weight so that deciduous and evergreen contribute symmetrically; weighted-F1 applies the empirical class proportions \(n_c/N\) observed in the fold, i.e.
\begin{equation*}
    \mathrm{F1}_{\text{macro}} = \frac{1}{2} \sum_{c \in \{\text{dec}, \text{ever}\}} \mathrm{F1}_c,
    \qquad
    \mathrm{F1}_{\text{weighted}} = \sum_{c \in \{\text{dec}, \text{ever}\}} \frac{n_c}{N} \mathrm{F1}_c,
\end{equation*}
where \(n_c\) is the number of validation pixels of class \(c\) and \(N = n_{\text{dec}} + n_{\text{ever}}\). This averaging prevents overall accuracy from masking poor minority-class performance, particularly for evergreen stands. We report the mean, standard deviation, and interquartile range across folds, while eco-region summaries average validation folds using the eco-region weights. Calibration is assessed with 10 equal-width probability bins on the evergreen posterior, i.e. the random forest probability that a pixel belongs to the evergreen class. For each bin \(b\) with sample set \(S_b\) we compute the mean confidence \(\hat{p}_b = |S_b|^{-1} \sum_{i \in S_b} p_i\) and the observed evergreen fraction \(\hat{y}_b = |S_b|^{-1} \sum_{i \in S_b} \mathbf{1}\{y_i = \text{evergreen}\}\). The expected calibration error is the weighted average of the absolute gaps between these two quantities,
\begin{equation*}
    \mathrm{ECE} = \sum_{b=1}^{10} \frac{|S_b|}{N} \left| \hat{p}_b - \hat{y}_b \right|,
\end{equation*}
with \(N\) the total number of validation pixels, while the maximum calibration error (MCE) records \(\max_b \left| \hat{p}_b - \hat{y}_b \right|\). Low ECE and MCE indicate that stated probabilities match observed frequencies, making the model's confidence trustworthy for threshold tuning. When a model is perfectly calibrated, \(\hat{p}_b\) and \(\hat{y}_b\) coincide in every bin and both metrics vanish. This reliability analysis clarifies how users can adjust the decision threshold: we also track macro-F1 when the evergreen threshold moves to 0.45 or 0.55 to confirm that modest shifts—such as raising the threshold to suppress evergreen false positives—do not destabilise overall performance.
